---
title: "README"
author: "Michael B. Sohn"
date: "01/20/2026"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## MIC: <ins>M</ins>ultiple <ins>I</ins>mputation Method for High-Sparse, High-Dimensional, <ins>C</ins>ompositional Data

The *mic* function generates a number of imputed compositional data using a random selection and amalgamation procedure. For detailed information about the arguments, please see the documentation for *mic()*.

### Installation

install.packages("devtools")

devtools::install_github("mbsohn/mic")


### Example: Generate 20 imputed profiles

```{r, message=FALSE, warning=FALSE}
library(mic); library(tidyverse); library(emdbook)
# Simulate datasets for two groups using zero-inflated negative binomial models
set.seed(2026)
n.smpl <- 50; n.taxa <- 100; prop.str.zeros <- 0.5; da.taxa.prop <- 0.2
n.low.abun <- round(n.taxa*0.6)
n.mid.abun <- round(n.taxa*0.3)
n.high.abun <- n.taxa - n.low.abun - n.mid.abun
sim.mean1 <- sim.mean2 <- c(sample(1:3, n.low.abun, rep=TRUE),
                            sample(4:10, n.mid.abun, rep=TRUE),
                            sample(11:100, n.high.abun, rep=TRUE))
sim.wt <- sample(1:20, size=n.smpl, rep=TRUE)
sim.size <- sample(seq(0.5, 2, 0.1), n.taxa, rep=TRUE)
M1 <- rzinbinom(n = n.smpl*n.taxa,
                mu = as.vector(outer(sim.wt, sim.mean1, "*")),
                size = rep(sim.size, each=n.smpl),
                zprob = prop.str.zeros) %>%
  matrix(., nrow=n.smpl, ncol=n.taxa)
da.indx <- sample(1:n.taxa, round(n.taxa*da.taxa.prop))
sim.mean2[da.indx] <- sim.mean1[da.indx]*sample(c(seq(0.1, 0.5, 0.05), seq(2, 10, 0.5)),
                                                round(n.taxa*da.taxa.prop), rep=T)
M2 <- rzinbinom(n = n.smpl*n.taxa,
                mu = as.vector(outer(sim.wt, sim.mean2, "*")),
                size = rep(sim.size, each=n.smpl),
                zprob = prop.str.zeros) %>%
  matrix(., nrow=n.smpl, ncol=n.taxa)
colnames(M1) <- colnames(M2) <- paste0("T", 1:n.taxa)
# Combine M1 and M2 to construct the final profile
M.f <- rbind(M1, M2)
rownames(M.f) <- paste0("S", 1:nrow(M.f))
# Assign group memberships
y <- c(rep(0, n.smpl), rep(1, n.smpl))
# Get differentially abundant taxa
is.true.da.taxa <- (sim.mean1 != sim.mean2)
true.da.taxa <- paste0("T", which(is.true.da.taxa))
# Run mic
rslt <- mic(M=M.f, G=y, nimp=20)
```

### Differential abundance analysis using OPTIMEM
```{r, message=FALSE, warning=FALSE}
library(optimem); library(sandwich)
# To identify a reference set for the additive log-ratio transformation
rslt.opt <- optimem(M=M.f, y=y, eta=0.001, early.exit=FALSE)
##### 
da.rslt.coef <- da.rslt.var <- da.rslt.pval <- 
  matrix(0, nrow=length(rslt$Group1), ncol=ncol(rslt$Group1[[1]]))
colnames(da.rslt.coef) <- colnames(da.rslt.var) <- 
  colnames(da.rslt.pval) <- colnames(rslt$Group1[[1]])
for(i in 1:length(rslt$Group1)){
  M.tmp <- rbind(rslt$Group1[[i]], rslt$Group2[[i]])
  refs.tmp <- rowSums(M.tmp[, rslt.opt$nonDAtaxa_min])
  alrM.tmp <- sweep(M.tmp, 1, refs.tmp, "/") %>% log()
  for(j in 1:ncol(alrM.tmp)){
    lm.tmp <- lm(alrM.tmp[,j] ~ y)
    da.rslt.coef[i,j] <- summary(lm.tmp)$coef[2, "Estimate"]
    da.rslt.var[i,j] <- vcovHC(lm.tmp, type="HC3")[2,2]
    da.rslt.pval[i,j] <- summary(lm.tmp)$coef[2, "Pr(>|t|)"]
  }
}
# Pool the estimates
da.rslt <- pool_mic(t(da.rslt.coef), t(da.rslt.var))
cat("DA taxa: ", true.da.taxa, "\n")
print(da.rslt)
```
